---
title: "R Markdown example for the Reproducible Research course by Coursera"
author: "Serge Shmygelskyy aka Shmyg"
date: "16.04.2015"
output: html_document
---

This document is created as per Assignment #2 of the course.

## Data loading

Normally the file is already present in the repository, but I prefer to check if it's there just in case. Workflow:

* look for the file in the current directory
* if not found, download
* decompress
* read into a dataframe

```{r, echo = TRUE}
url <- 'https://d396qusza40orc.cloudfront.net/repstats%2Fstats%2Factivity.zip'
zipfile <- 'activity.zip'

if (!file.exists(zipfile)) {
        download.file (url, zipfile, method = 'curl')
} 
unzip(zipfile)
stats <- read.csv("activity.csv")

```

## Number of steps taken per day, histogram, mean and median values

```{r, echo=TRUE}
library('dplyr')
days <- group_by (stats, date)
finalData2 <- summarize(days, steps = sum(steps))
hist(finalData2$steps,
        main = 'Histogram of total number of steps per day',
        xlab = 'Number of steps')

mean(finalData2$steps, na.rm = TRUE)
median(finalData2$steps, na.rm = TRUE)

```

## Average daily activity pattern

Calculated by means of standard SQL
```{r, echo=TRUE}
library('sqldf')
avgPerInterval <- sqldf('select interval, avg(steps) as steps from stats group by interval')
plot (avgPerInterval, type = 'h')
avgPerInterval[which.max(avgPerInterval$steps),]
```

So the the winner (5-minute interval with the maximum number of steps) is `r avgPerInterval[which.max(avgPerInterval$steps),]$interval`

## Imputing missing data

Here we'll replace missing data (_steps_ variable) with the mean value of steps taken during this interval accross all the measurements

Number of incomplete cases:

```{r, echo=TRUE}
sum(is.na(stats$steps))
```

Imputing data

```{r, echo=TRUE}
completeData <- stats
for (i in 1:nrow(completeData)) { 
        if ( is.na(completeData[i,]$steps )) {
                completeData[i,]$steps <- avgPerInterval[avgPerInterval$interval == completeData[i,]$interval, "steps"]
        }
}

days <- group_by (completeData, date)
finalData4 <- summarize(days, steps = sum(steps))

hist(finalData4$steps,
        main = 'Histogram of total number of steps per day with imputed values',
        xlab = 'Number of steps')

mean(finalData4$steps)
median(finalData4$steps)
```

There are no tangible changes in the histogram after the missing values have been removed

## Difference in activity patterns between weekdays adn weekends

Here we'll add a column to the dataframe using _weekdays()_ function.

```{r, echo=TRUE}
for (i in 1:nrow(completeData)) {
        day <- weekdays(as.Date(completeData[i, "date"]))
        if (day %in% c('Saturday', 'Sunday')) {
                completeData[i, "day"] <- 'Weekend'
        }
        else {
                completeData[i, "day"] <- 'Weekday'
        }
}

completeData <- transform(completeData, day = factor(day))
dayTypes <- group_by (completeData, day)
finalData5 <- summarize(dayTypes, steps = sum(steps))
```